Linear and Differential Cryptanalysis of DES

### 2009-02-02 talk/draft1.tex
Now there is the final final version ... :)

\vspace{2mm}
\pause
\begin{itemize}[<+->]
\item{How many pairs do we need?}
\item{Approximately $2^4 = 16$ for $100\%$ success}
\end{itemize}

@@

\begin{frame}
The problems of DES originated in two aspects:
\begin{itemize}
\item{6 bit S-Boxes are too easy to identify with linear or affine functions with high bias}
\item{Diffusion between rounds is just not fast enough}
\end{itemize}
AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
\end{frame}

\begin{frame}
Matsui rearrangement of S-Boxes ... Construction of S-Boxes ... IBM knew Diff 17 years earlier ... $10^44$ is closer to $10$ than to $10^45$ ...
\end{frame}

>>>>

\vspace{2mm}
\pause
\begin{itemize}[<+->]
\item{Question: How many pairs do we need?}
\item{Answer: $N \approx c/p_D$ where $p_D$ is the probability of the differential characteristic, and $c > 0$ is a constant}
\end{itemize}
\pause I used $2^4 = 16$ pairs for $100\%$ success in this case. For 16 round DES, $2^{47}$ pairs are needed, out of which $2^{36}$ are ``good pairs''.
\end{frame}

@@

\begin{frame}
If you are designing a block cipher, note:

\vspace{5mm}
\pause The problems of DES originated in two aspects
\begin{itemize}[<+->]
\item{6 bit S-Boxes are too easy to identify with linear or affine functions with high bias}
\item{Diffusion between rounds is just not fast enough}
\end{itemize}

\vspace{2mm}
\pause Prescription
\begin{itemize}[<+->]
\item{Create S-Boxes as bent functions [inefficient as $p \geq 2q$]}
\item{Make the diffusion step faster to create an avalanche}
\end{itemize}

\vspace{3mm}
\pause AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
\end{frame}

\subsection{Fun facts}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Just for Fun!
\end{beamercolorbox}
\end{frame}

\begin{frame}
What if you play around with the S-Box ordering?

\vspace{5mm}
\pause
\begin{itemize}[<+->]
\item{Matsui prescribed a rearrangement of S-Boxes specifically against linear attack.}
\item{Matsui prescribed another rearrangement of the S-Boxes, which is good enough to prevent both linear and differential attacks.}
\item{Biham and Shamir illustrated a rearrangement where differential attack becomes way more efficient.}
\end{itemize}

\begin{frame}
\begin{figure}
\includegraphics[width=0.6\textwidth]{ibm.jpg}
\end{figure}

\pause
S-Box Design criteria
\begin{itemize}[<+->]
\item{1 bit input difference produces 2 bits output difference}
\item{Minimize the difference between the numbers of 1's and 0's when any input bit is held constant}
\item{$S(X) \neq S(X \oplus 11ab00)$} 
\end{itemize}

\vspace{2mm}
\pause Other good stuff
\begin{itemize}[<+->]
\item{2 input difference bits mapped to 3 by $E(R)$}
\item{Each output difference bit permuted to 1 each to S-boxes of next round}
\end{itemize}

\begin{frame}
\begin{figure}
\includegraphics[totalheight=0.7\textheight]{coppersmith.jpg}
\end{figure}
\pause
\centerline{Average of $10$ and $10^{45}$ is $5.10^{44} + 5 > 10^{44} \approx 2^{47}$}
\pause
\begin{figure}
\includegraphics[width=0.7\textwidth]{eli.jpg}
\end{figure}
\hfill - Eli Biham
\end{frame}
